Estimation below the detection limit with application to cryo-em

### 2018-08-08 TEX/inverse_problem_draft.tex

\newcommand{\LL}{\mathcal{L}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\ep}{\varepsilon}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\GCD}{\mathbf{GCD}}
\newcommand{\XX}{\mathcal{X}}
\newcommand{\SUM}{\text{sum}}
\newcommand{\1}{\mathbf{1}}

@@

\newcommand{\reals}{\mathbb{R}}
\newcommand{\RL}{\mathbb{R}^L}
\newcommand{\CL}{\mathbb{C}^L}
\newcommand{\RN}{\mathbb{R}^N}
\newcommand{\RNN}{\mathbb{R}^{N\times N}}
\newcommand{\CNN}{\mathbb{C}^{N\times N}}
\newcommand{\inner}[1]{\left\langle {#1} \right\rangle}
\newcommand{\hx}{\hat{x}} 
\newcommand{\one}{\mathbf{1}} 
\newcommand{\SNR}{\ensuremath{\textsf{SNR}}}

@@

Let $x_1,\ldots,x_K\in\RL$ be the sought signals and let $y\in\RN$ be the observed micrograph (notice that it is equivalent to think of the data as being one long micrograph or multiple shorter micrographs concatenated into one.)
The forward model (or ``image'' formation model) is as follows. For each target signal $x_k$, an unknown binary signal $s_k \in \{0, 1\}^{N-L+1}$ indicates (with 1's) the starting positions of all occurrences of $x_k$ in $y$, so that, with additive white Gaussian noise:
\begin{align}
y & = \sum_{k=1}^K x_k \ast s_k + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2 I_N),
\label{eq:model}
\end{align}
where $\ast$ denotes linear convolution. The binary signals obey the following property:
\begin{align}
\textrm{If } s_k[i] = 1 \textrm{ and } s_{k'}[j] = 1 \textrm{ but } (k, i) \neq (k', j), \textrm{ then } |i - j| \geq 2L-1.
\label{eq:spacing}
\end{align}
In words: the starting positions of any two occurrences (be it of the same signal or of two different signals) must be separated by at least $2L-1$ positions, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the micrograph.
It can be also extended to Poison model; see Section TKTK.

@@

\subsection{Poisson model (rename)}

@@

Having computed the moments of interest, we now estimate signals $x_1, \ldots, x_K$ and coefficients $\gamma_1, \ldots, \gamma_K$ which agree with the data. We choose to do so by running an optimization algorithm on the following nonlinear least-squares problem:
\begin{multline}
\min_{\substack{\hat x_1, \ldots, \hat x_K \in \reals^{W} \\ \hat \gamma_1, \ldots, \hat \gamma_K > 0}} w_1 \left( a_y^1 - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^1 \right)^2 + w_2 \sum_{\ell = 1}^{L-1} \left( a_y^2[\ell] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^2[\ell] \right)^2 + \\ w_3 \sum_{\substack{2\leq\ell_1\leq L-1 \\ 1 \leq \ell_2 \leq \ell_1-1}} \left( a_y^3[\ell_1, \ell_2] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^3[\ell_1,\ell_2] \right)^2,
\label{eq:optim1D}
\end{multline}
where $W \geq L$ is the length of the sought signals and the weights are set to $w_1 = 1/2, w_2 = 1/2n_2, w_3 = 1/2n_3$, where $n_2, n_3$ are the number of moments used: $n_2 = L-1$, $n_3 = \frac{(L-1)(L-2)}{2}$ (weights could also be set in accordance with variance estimates as in~\cite{boumal2017heterogeneous}).

@@

>>>>

\newcommand{\LL}{\mathcal{L}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\ep}{\varepsilon}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\GCD}{\mathbf{GCD}}
\newcommand{\XX}{\mathcal{X}}
\newcommand{\SUM}{\text{sum}}
\newcommand{\1}{\mathbf{1}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Poisson}{\text{Poisson}}
%\renewcommand{\P}{\mathbb{P}}
%\renewcommand{\L}{\mathcal{L}}

@@

\newcommand{\R}{\mathbb{R}}

\newcommand{\RL}{\mathbb{R}^L}
\newcommand{\CL}{\mathbb{C}^L}
\newcommand{\RN}{\mathbb{R}^N}
\newcommand{\RNN}{\mathbb{R}^{N\times N}}
\newcommand{\CNN}{\mathbb{C}^{N\times N}}
\newcommand{\inner}[1]{\left\langle {#1} \right\rangle}
\newcommand{\hx}{\hat{x}} 
\newcommand{\one}{\mathbf{1}} 
\newcommand{\SNR}{\ensuremath{\textsf{SNR}}}

@@

Let $x_1,\ldots,x_K\in\RL$ be the sought signals and let $y\in\RN$ be the observed micrograph (notice that it is equivalent to think of the data as being one long micrograph or multiple shorter micrographs concatenated into one.)
The forward model (or ``image'' formation model) is as follows. For each target signal $x_k$, an unknown binary signal $s_k \in \{0, 1\}^{N-L+1}$ indicates (with 1's) the starting positions of all occurrences of $x_k$ in $y$, so that, with additive white Gaussian noise:
\begin{align}
y & = \sum_{k=1}^K x_k \ast s_k + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2 I_N),
\label{eq:model}
\end{align}
where $\ast$ denotes linear convolution. 
We consider two models for the generative process of the sparse signal:

\paragraph{The well separated support model.}
The binary signals obey the following property:
\begin{align}
\textrm{If } s_k[i] = 1 \textrm{ and } s_{k'}[j] = 1 \textrm{ but } (k, i) \neq (k', j), \textrm{ then } |i - j| \geq 2L-1.
\label{eq:spacing}
\end{align}
In words: the starting positions of any two occurrences (be it of the same signal or of two different signals) must be separated by at least $2L-1$ positions, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the micrograph.

\paragraph{Poisson process model.} In this model, points are chosen in $\{1,\dots,n\}$ according to a Poisson process with parameter $\gamma n$. For each point $i$ that is chosen from $1$ to $n$, a random vector $X$ from the distribution is then placed in the large vector, with element $0$ at location $i$, with overlapping vectors being added together.

If $M_i$ denotes the number of hits at location $i$, $1 \le i \le n$, then by definition of the Poission process $M_i$'s are i.i.d.\
 and $M_i \sim \Poisson(\gamma)$. Conditional on the value of $M = (M_1,\dots,M_n)$, if we let $X_1^{i},\dots,X_{M_i}^i$ denote the random vectors with position 0 located at $i$, then $X_{k_1}^{i}$ and $X_{k_2}^{i}$ are independent for $k_1 \ne k_2$.

With this notation, if $Y \in \R^{n+L}$ is the observed vector, we can write each entry as:
%
\begin{align}
%
Y[i] = \sum_{j=0}^{L-1} \sum_{k=1}^{M_{i-j}} X_k^{i-j}[j].
%
\end{align}

@@

\subsection{Autocorrelations of the Poisson process model}

From the first three autocorrelations of the Poisson model one can compute the first three autocorrelations of the well-separated model.

\begin{proposition} \label{prop:poisson_model}
Under the Poisson process model, the first autocorrelation  is $\gamma \LL_1$, the second moment vector is $(\gamma \LL_1)^2 + \gamma \LL_2(\Delta)$, and the third moment matrix is $(\gamma \LL_1)^3 + \gamma \LL_1  \cdot ( \gamma\LL_2(\Delta_1) + \gamma\LL_2(\Delta_2) + \gamma\LL_2(\Delta_2-\Delta_1)) + \gamma \LL_3(\Delta_1,\Delta_2) $. In particular, from the first three moments of the Poisson process model, once can recover the first three moments from the strongly-separated model, with the Poisson rate $\gamma$ playing the role of the ``occupancy factor''. So if recovery is possible for the strongly-separated model, it is also possible for the Poisson process model.
\end{proposition}

@@

Having computed the moments of interest, we now estimate signals $x_1, \ldots, x_K$ and coefficients $\gamma_1, \ldots, \gamma_K$ which agree with the data. We choose to do so by running an optimization algorithm on the following nonlinear least-squares problem:
\begin{multline}
\min_{\substack{\hat x_1, \ldots, \hat x_K \in \R^{W} \\ \hat \gamma_1, \ldots, \hat \gamma_K > 0}} w_1 \left( a_y^1 - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^1 \right)^2 + w_2 \sum_{\ell = 1}^{L-1} \left( a_y^2[\ell] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^2[\ell] \right)^2 + \\ w_3 \sum_{\substack{2\leq\ell_1\leq L-1 \\ 1 \leq \ell_2 \leq \ell_1-1}} \left( a_y^3[\ell_1, \ell_2] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^3[\ell_1,\ell_2] \right)^2,
\label{eq:optim1D}
\end{multline}
where $W \geq L$ is the length of the sought signals and the weights are set to $w_1 = 1/2, w_2 = 1/2n_2, w_3 = 1/2n_3$, where $n_2, n_3$ are the number of moments used: $n_2 = L-1$, $n_3 = \frac{(L-1)(L-2)}{2}$ (weights could also be set in accordance with variance estimates as in~\cite{boumal2017heterogeneous}).

@@

\section{Proof of Proposition~\ref{prop:poisson_model}}

%\section{Setup}
%
%We consider the following observation model. $X = (X[0],\dots,X[L-1])$ is a random vector of length $L$, drawn from some fixed distribution. For fixed $n$, we observe a random vector $Y$ of length $n+L$, generated as follows. Points are chosen in $\{1,\dots,n\}$ according to a Poisson process with parameter $\gamma n$. For each point $i$ that is chosen from $1$ to $n$, a random vector $X$ from the distribution is then placed in the large vector, with element $0$ at location $i$, with overlapping vectors being added together. 
%
%If $M_i$ denotes the number of hits at location $i$, $1 \le i \le n$, then by definition of the Poission process $M_i$'s are iid and $M_i \sim \Poisson(\gamma)$. Conditional on the value of $M = (M_1,\dots,M_n)$, if we let $X_1^{i},\dots,X_{M_i}^i$ denote the random vectors with position 0 located at $i$, then $X_{k_1}^{i}$ and $X_{k_2}^{i}$ are independent for $k_1 \ne k_2$.
%
%With this notation, if $Y \in \R^{n+L}$ is the observed vector, we can write each entry as:
%%
%\begin{align}
%%
%Y[i] = \sum_{j=0}^{L-1} \sum_{k=1}^{M_{i-j}} X_k^{i-j}[j].
%%
%\end{align}

We will denote by $\M_l$ the moments of $X$:
%
\begin{align}
%
\M_1[i] = \E X[i], \quad 0 \le i \le L-1,
%
\end{align}
%
\begin{align}
%
\M_2[i,j] = \E X[i] X[j], \quad 0 \le i,j \le L-1,
%
\end{align}
%
and
%
\begin{align}
%
\M_3[i,j,k] = \E X[i] X[j] X[k], \quad 0 \le i,j,k \le L-1.
%
\end{align}

We will also denote by $\L_l$ the autocorrelations of $X$:
%
\begin{align}
%
\LL_1 = \sum_{i=0}^{L-1} \M_1[i],
%
\end{align}
%
\begin{align}
%
\LL_2(\Delta) = \sum_{i=0}^{L-1} \M_2[i,i+\Delta],
%
\end{align}
%
and
%
\begin{align}
%
\LL_3(\Delta_1,\Delta_2) = \sum_{i=0}^{L-1} \M_3[i,i+\Delta_1,i+\Delta_2].
%
\end{align}

Note that in the strongly-separated model, the first three observed moments are, respectively, $\LL_1$, $\LL_2(\Delta)$, and $\LL_3(\Delta_1,\Delta_2)$.

%

In this notation, we will show that the first moment of the data is $\gamma \LL_1$, the second moment vector is $(\gamma \LL_1)^2 + \gamma \LL_2(\Delta)$, and the third moment matrix is $(\gamma \LL_1)^3 + \gamma \LL_1  \cdot ( \gamma\LL_2(\Delta_1) + \gamma\LL_2(\Delta_2) + \gamma\LL_2(\Delta_2-\Delta_1)) + \gamma \LL_3(\Delta_1,\Delta_2) $. In particular, from the first three moments of the Poisson process model, once can recover the first three moments from the strongly-separated model, with the Poisson rate $\gamma$ playing the role of the ``occupancy factor''. So if recovery is possible for the strongly-separated model, it is also possible for the Poisson process model.

%

\section{The first moment of $Y$}

To compute the first moment of $Y$, we will first condition on $M = (M_1,\dots,M_n)$, and then average over $M$. We have:
%
\begin{align}
%
\E[Y[i] | M] = \sum_{j=0}^{L-1} \sum_{k=1}^{M_{i-j}} \E X_k^{i-j}[j]
= \sum_{j=0}^{L-1} \sum_{k=1}^{M_{i-j}} \M_1[j]
= M_{i-j} \sum_{j=0}^{L-1} \M_1[j].
%
\end{align}
%
Now taking expectations over $M$ we see:
%
\begin{align}
%
\E Y[i] = \gamma \sum_{j=0}^{L-1} \M_1[j] = \gamma \LL_1.
%
\end{align}

%

\section{The second moment of $Y$}

Again, we will condition on $M$ first, and then take the expectation over $M$. Fix $i_1 \ne i_2$, and let $\Delta = i_2 - i_1$. Then:
%
\begin{align}
%
Y_{i_1} Y_{i_2} 
&= \sum_{j_1=0}^{L-1} \sum_{j_2=0}^{L-1} 
\sum_{k_1=1}^{M_{i_1-j_1}}\sum_{k_2=1}^{M_{i_2-j_2}}
X_{k_1}^{i_1-j_1}[j_1] X_{k_2}^{i_2 - j_2}[j_2].
%
\end{align}

We break up the double sum over $j_1$ and $j_2$ into two terms: one where $j_2 \ne j_1 + \Delta$, and one where $j_2 = j_1 + \Delta$ or equivalently $i_1-j_1 = i_2-j_2$. In the first case, all the terms are independent, and so the expectation factors. In the second case, when $k_1 \ne k_2$ we have independence, but otherwise not. This gives (all expectations are conditional on $M$):

\begin{align}
%
\E Y_{i_1} Y_{i_2}
=& \sum_{j_1=0}^{L-1} \sum_{j_2=0}^{L-1} 
\sum_{k_1=1}^{M_{i_1-j_1}}\sum_{k_2=1}^{M_{i_2-j_2}}
\E X_{k_1}^{i_1-j_1}[j_1] X_{k_2}^{i_2 - j_2}[j_2]
\nonumber \\
=& \sum_{j_1 - j_2 \ne \Delta} \sum_{k_1} 
\sum_{k_2} \E X_{k_1}^{i_1-j_1}[j_1] X_{k_2}^{i_2 - j_2}[j_2]
\nonumber \\
& + \sum_{j_1 = 0}^{L-1} \sum_{k_1 \ne k_2} 
\E X_{k_1}^{i_1-j_1}[j_1] X_{k_2}^{i_1 - j_1}[j_1+\Delta]
\nonumber \\
& + \sum_{j_1 = 0}^{L-1} \sum_{k_1=1}^{M_{i_1-j_1}} 
\E X_{k_1}^{i_1-j_1}[j_1] X_{k_1}^{i_1-j_1}[j_1 + \Delta] 
\nonumber \\
=& \sum_{j_1 - j_2 \ne \Delta} M_{i_1-j_1} M_{i_2 - j_2} \M_1[j_1] \M_1[j_2]
\nonumber \\
& + \sum_{j_1 = 0}^{L-1} M_{i_1-j_1}(M_{i_1-j_1} - 1) \M_1[j_1] \M_1[j_1 + \Delta]
\nonumber \\
& + \sum_{j_1 = 0}^{L-1} M_{i_1-j_1} \M_2[j_1,j_1 + \Delta] .
%
\end{align}
%
Now take expectations over the Poisson random variables, using this fact:
%
\begin{lem} \label{lem-choose}
	If $M \sim \Poisson(\gamma)$, then 
	\begin{align}
	%
	\E {M\choose k} = \frac{\gamma^k}{k!}.
	%
	\end{align}
\end{lem}

We get (now the expectation is over $M$ and $X$):
%
\begin{align}
%
\E Y_{i_1} Y_{i_2} 
=& \sum_{j_1 - j_2 \ne \Delta} \E M_{i_1-j_1} M_{i_2 - j_2} \M_1[j_1] \M_1[j_2]
\nonumber \\
& + \sum_{j_1 = 0}^{L-1} \E M_{i_1-j_1}(M_{i_1-j_1} - 1) \M_1[j_1] \M_1[j_1 + \Delta]
\nonumber \\
& + \sum_{j_1 = 0}^{L-1} \E M_{i_1-j_1} \M_2[j_1,j_1 + \Delta]
\nonumber \\
=& \sum_{j_1 - j_2 \ne \Delta} \gamma^2 \M_1[j_1] \M_1[j_2]
+ \sum_{j_1 = 0}^{L-1} \gamma^2 \M_1[j_1] \M_1[j_1 + \Delta]
\nonumber \\
& + \sum_{j_1 = 0}^{L-1} \gamma \M_2[j_1,j_1 + \Delta]
\nonumber \\
=&  \bigg(\gamma \sum_{j = 0}^{L-1} \M_1[j] \bigg)^2
+ \gamma \sum_{j = 0}^{L-1} \M_2[j,j + \Delta]
\nonumber \\
=&  (\gamma \LL_1)^2 + \gamma \LL_2(\Delta).
%
\end{align}

But the first term in the sum is just the square of the first moment of $Y$; so from the first two moments we can recover $\gamma \LL_2(\Delta)$, which is just the expected power spectrum of the random vector $X$, i.e.\ the usual second moment we have been working with.

%

\section{The third moment of $Y$}

For three distinct $i_1$, $i_2$ and $i_3$, we let $\Delta_1 = i_2 - i_1$ and $\Delta_2 = i_3 - i_1$. We have:
%
\begin{align}
%
&Y_{i_1} Y_{i_2} Y_{i_3}
\nonumber \\
&= \sum_{j_1=0}^{L-1} \sum_{j_2=0}^{L-1} \sum_{j_3=0}^{L-1} 
\sum_{k_1=1}^{M_{i_1-j_1}}\sum_{k_2=1}^{M_{i_2-j_2}} \sum_{k_3=1}^{M_{i_3-j_3}}
X_{k_1}^{i_1-j_1}[j_1] X_{k_2}^{i_2 - j_2}[j_2] X_{k_3}^{i_3 - j_3}[j_3].
\end{align}
%
We will break up the outer three sums into disjoint sums with the following ranges of indices:
%
\begin{enumerate}

	\item \label{case1}
	$j_2 = j_1 + \Delta_1$ and $j_3 = j_2 + \Delta_2 - \Delta_1$.

	\item \label{case2}
	$j_2 = j_1 + \Delta_1$ and $j_3 \ne j_2 + \Delta_2 - \Delta_1$.

	\item \label{case3}
	$j_2 \ne j_1 + \Delta_1$ and $j_3 = j_1 + \Delta_2$.

	\item \label{case4}
	$j_2 \ne j_1 + \Delta_1$ and $j_3 \ne j_1 + \Delta_2$ and $j_3 = j_2 + \Delta_2 - \Delta_1$.

	\item \label{case5}
	$j_2 \ne j_1 + \Delta_1$ and $j_3 \ne j_1 + \Delta_2$ and $j_3 \ne j_2 + \Delta_2 - \Delta_1$.

\end{enumerate}

For Case \ref{case1}, we have $\ell \equiv i_1 - j_1 = i_2 - j_2 = i_3 - j_3$. We further break up the sum:
%
\begin{align}
%
&\sum_{j=0}^{L-1} \sum_{k_1=1}^{M_\ell} \sum_{k_2=1}^{M_\ell} \sum_{k_3=1}^{M_\ell} 
X_{k_1}^{\ell}[j] X_{k_2}^{\ell}[j + \Delta_1] X_{k_3}^{\ell}[j + \Delta_2]
\nonumber \\
=& \underbrace{ \sum_{j=0}^{L-1} \sum_{k_i \text{distinct}} 
	X_{k_1}^{\ell}[j] X_{k_2}^{\ell}[j + \Delta_1] X_{k_3}^{\ell}[j + \Delta_2]
}_{\text{(a)}}
\nonumber \\
&+\underbrace{ \sum_{j=0}^{L-1} \sum_{k_1=k_2\ne k_3} 
	X_{k_1}^{\ell}[j] X_{k_2}^{\ell}[j + \Delta_1] X_{k_3}^{\ell}[j + \Delta_2]
}_{\text{(b)}}
\nonumber \\
&+\underbrace{ \sum_{j=0}^{L-1} \sum_{k_1=k_3\ne k_2} 
	X_{k_1}^{\ell}[j] X_{k_2}^{\ell}[j + \Delta_1] X_{k_3}^{\ell}[j + \Delta_2]
}_{\text{(c)}}
\nonumber \\
&+\underbrace{ \sum_{j=0}^{L-1} \sum_{k_2=k_3\ne k_1} 
	X_{k_1}^{\ell}[j] X_{k_2}^{\ell}[j + \Delta_1] X_{k_3}^{\ell}[j + \Delta_2]
}_{\text{(d)}}
\nonumber \\
&+\underbrace{ \sum_{j=0}^{L-1} \sum_{k_1=k_2=k_3} 
	X_{k_1}^{\ell}[j] X_{k_2}^{\ell}[j + \Delta_1] X_{k_3}^{\ell}[j + \Delta_2]
}_{\text{(e)}}.
%
\end{align}

For term (a), the expectation conditional on $M$ is:
%
\begin{align}
%
\sum_{j=0}^{L-1} M_\ell(M_\ell-1)(M_\ell-2)\M[j] \M[j+\Delta_1] \M[j+\Delta_2].
%
\end{align}
%
Using Lemma \ref{lem-choose}, the unconditional expectation of (a) is then:
%
\begin{align} \label{aaaa}
%
\gamma^3 \sum_{j=0}^{L-1} \M_1[j] \M_1[j+\Delta_1] \M_1[j+\Delta_2].
%
\end{align}

For term (b), the expectation conditional on $M$ is:
%
\begin{align}
%
\sum_{j=0}^{L-1} M_\ell (M_\ell - 1) \M_2[j,j+\Delta_1] \M_1[j + \Delta_2]
%
\end{align}
%
and then again using Lemma \ref{lem-choose} we get the expected value:
%
\begin{align} \label{bbbb}
%
\gamma^2 \sum_{j=0}^{L-1} \M_2[j,j+\Delta_1] \M_1[j+\Delta_2].
%
\end{align}

Similarly, the expected values of terms (c) and (d) are:
%
\begin{align} \label{cccc}
%
\gamma^2 \sum_{j=0}^{L-1} \M_2[j,j+\Delta_2] \M_1[j+\Delta_1].
%
\end{align}
%
and
%
\begin{align} \label{dddd}
%
\gamma^2 \sum_{j=0}^{L-1} \M_2[j+\Delta_1,j+\Delta_2] \M_1[j].
%
\end{align}

Finally, the expected value of term (e) is easily shown to be:
%
\begin{align} \label{eeee}
%
\gamma \sum_{j=0}^{L-1} \M_3[j,j+\Delta_1,j+\Delta_2].
%
\end{align}
%
This concludes the computation for Case \ref{case1}.

Moving onto Case \ref{case2}, we have $\ell_1 \equiv i_1 - j_1 = i_2 - j_2$, and also define $\ell_2 \equiv i_3 - j_3$. By definition, $\ell_1 \ne \ell_2$. The sum is:
%
\begin{align}
%
& \sum_{j_1=0}^{L-1} \sum_{j_3 \ne j_1 + \Delta_2}
\sum_{1 \le k_1,k_2 \le M_{\ell_1}} \sum_{k_3=1}^{M_{\ell_2}}
X_{k_1}^{\ell_1}[j_1] X_{k_2}^{\ell_1}[j_1 + \Delta_1] X_{k_3}^{\ell_2}[j_3]   
\nonumber \\
=& \sum_{j_1=0}^{L-1} \sum_{j_3 \ne j_1 + \Delta_2} \sum_{k_3=1}^{M_{\ell_2}}
\Bigg\{
\sum_{1 \le k_1 \ne k_2 \le M_{\ell_1}} 
X_{k_1}^{\ell_1}[j_1] X_{k_2}^{\ell_1}[j_1 + \Delta_1] X_{k_3}^{\ell_2}[j_3]
\nonumber \\
&       + \sum_{k_1=1}^{M_{\ell_1}} X_{k_1}^{\ell_1}[j_1] X_{k_1}^{\ell_1}[j_1 + \Delta_1] 
X_{k_3}^{\ell_2}[j_3]  \Bigg\}.
%
\end{align}
%
Taking expectations conditional on $M$, we then get:
%
\begin{align}
%
& \sum_{j_1=0}^{L-1} \sum_{j_3 \ne j_1 + \Delta_2} 
\Bigg(M_{\ell_1} (M_{\ell_1}-1) M_{\ell_2} \M_1[j_1] \M_1[j_1 + \Delta_1] \M_1[j_3]
\nonumber \\
&       + M_{\ell_1} M_{\ell_2} \M_2[j_1,j_1+\Delta_1] \M_1[j_3] \Bigg).
%
\end{align}

Taking expectations over $M$ and using Lemma \ref{lem-choose} then gives:
%
\begin{align}
%
& \gamma^3 \sum_{j_1=0}^{L-1} \sum_{j_3 \ne j_1 + \Delta_2} 
\M_1[j_1] \M_1[j_1 + \Delta_1] \M_1[j_3]
\label{ffff} \\
& + \gamma^2 \sum_{j_1=0}^{L-1} \sum_{j_3 \ne j_1 + \Delta_2}  
\M_2[j_1,j_1+\Delta_1] \M_1[j_3].
\label{gggg}
%
\end{align}

Similarly, Cases \ref{case3} and \ref{case4} give the expressions:
%
\begin{align}
%
& \gamma^3 \sum_{j_1=0}^{L-1} \sum_{j_2 \ne j_1 + \Delta_1} 
\M_1[j_1] \M_1[j_1 + \Delta_2] \M_1[j_2]
\label{hhhh} \\
& + \gamma^2 \sum_{j_1=0}^{L-1} \sum_{j_2 \ne j_1 + \Delta_1}  
\M_2[j_1,j_1+\Delta_2] \M_1[j_2]
\label{iiii}
%
\end{align}
%
and
%
\begin{align}
%
& \gamma^3 \sum_{j_2=0}^{L-1} \sum_{j_1 \ne j_2} 
\M_1[j_1] \M_1[j_2 + \Delta_1] \M_1[j_2 + \Delta_2]
\label{jjjj}\\
& + \gamma^2 \sum_{j_2=0}^{L-1} \sum_{j_1 \ne j_2}  
\M_2[j_2+\Delta_1,j_2+\Delta_2] \M_1[j_1].
\label{kkkk}
%
\end{align}

Finally, in Case \ref{case5} we have $i_1 - j_1$, $i_2 - j_2$, and $i_3 - j_3$ are all pairwise distinct. Consequently, the $X$ variables are always independent, and the expectation conditional on $M$ (letting $\ell_q = i_q - j_q$, $q=1,2,3$),
%
\begin{align}
%
\sum_{j_1,j_2,j_3} M_{\ell_1} M_{\ell_2} M_{\ell_3} \M_1[j_1] \M_1[j_2] \M_1[j_3];
%
\end{align}
%
since the $M_{\ell_q}$'s are pairwise independent, $q=1,2,3$, the expectation over $M$ then yields:
%
\begin{align} \label{llll}
%
\gamma^3 \sum_{j_1,j_2,j_3} \M_1[j_1] \M_1[j_2] \M_1[j_3].
%
\end{align}

Now we add all the terms from Cases \ref{case1} to \ref{case5}. Expressions \eqref{aaaa}, \eqref{ffff}, \eqref{hhhh}, \eqref{jjjj}, and \eqref{llll} sum to the expression:
%
\begin{align}
%
(\gamma \LL_1)^3.
%
\end{align}

Note that this is obtained directly from the first moment. Expressions \eqref{bbbb}, \eqref{cccc}, \eqref{dddd}, \eqref{gggg},\eqref{iiii}, and \eqref{kkkk} sum to the expression:
%
\begin{align}
%
\gamma \LL_1  \cdot 
( \gamma\LL_2(\Delta_1) + \gamma\LL_2(\Delta_2) + \gamma\LL_2(\Delta_2-\Delta_1)).
%
\end{align}
%
Again, note that this is obtained directly from the first two moments. Finally, expression \eqref{eeee} is simply:
%
\begin{align}
%
\gamma \LL_3(\Delta_1,\Delta_2)
%
\end{align}
%
which is the usual third-order auto-correlation.

%

\section{Signal plus noise}

The expected values of the non-zero (for second moment) and off-diagonal (for third moment) terms are the same as without noise, as is true for the strongly-separated case. The same proof of almost sure convergence from my notes for the strongly-separated case also goes through verbatim.
